---
title: "Data structures"
author: "Hicham Zmarrou"
date: "March 15, 2017"
output: html_document
---


  
  
## Vectors
Numeric Vectors

There are several ways to assign values to a variable:
  

```{r}
a <- 1.7               # assign a value to a vector with only one element (~ scalar)
1.7 -> a               # assign a value to a vector with only one element (~ scalar)
a = 1.7                # assign a value to a vector with only one element (~ scalar)
assign("a", 1.7)       # assign a value to a vector with only one element (~ scalar)
```
To show the values:

```{r}  
 a                      # show the value in the screen (not valid in scripts)
print(a)               # show the value in the screen (valid in scripts)
```

To generate a vector with several numeric values:

```{r}  
a <- c(10, 11, 15, 19) # assign four values to a vector using the concatenate command c()
a                      # show the value in the screen
```
The operations are always done over all the elements of the numeric array:

```{r}  
a*a                    # evaluate the square value of every element in the vector
```

```{r}
1/a                    # evaluate the inverse value of every element in the vector
````

```{r}
b <- a-1               # subtract 1 from every element and assign the result to b
b
```

To generate a sequence:
```{r}  
2:10                                # generate a sequence from n1=2 to n2=10 using n1:n2
5:1                                 # generate an inverse sequence if n2 < n1
```

```{r}
# seq(from=n1, to=n2, by=n3)          # generate sequence from n1 to n2 using n3 step
seq(from=1, to=10, by=3)           #  (parameters names can be avoided if order is kept)
seq(from=1, to=10, by=3)
seq(1, 10, 3)
seq(length=10, from=1, by=3)        # generate a fixed length sequence
help(seq)                           # for help about this command
```

To generate repetitions:

```{r}  
a <- 1:3; b <- rep(a, times=3); c <- rep(a, each=3)         # command rep()
```

In the previous example we have run three commands in the same line. They have been separated by a ‘;’.

The content of the three variables is now:
```{r}  
a  # print(a)
b 
c
```
The recycling rule: vectors of different sizes can be combined, as far as the length of the longer vector is a multiple of the shorter vector’s length (otherwise a warning is issued, although the operation is carried out):

```{r}  
a+c                                         # proper dimensions
d <- c(10,100)
b+d                                         # incorrect dimensions
```

```{r}
ls()
rm(a,c)                                     # remove objects 'a' and 'b'
ls()                                        # list current objects
```

In order to remove everything in the working environment:
```{r}  
rm(list=ls())                               # Use this with caution
ls()                                        # (you'll receive no warning!)
```


```{r}
a <- seq(1:10)                                # generate a sequence
a                                           # show values in screen    
```
```{r}
b <- (a>5)                                  # assign values from an inequality
b                                           # show values in screen
```

```{r}
a[b]                                        # show values that fulfil the condition
```

```{r}
a[a>5]                                      # the same, but avoiding intermediate variable
```

```{r}
a <- "This is an example"                   # generate a character vector
a                                           # show vector content
``` 


```{r}
x <- 1.5
y <- -2.7
paste("Point is (",x,",",y,")", sep="")     # concatenate x, y and a string using 'paste'

```

### Matrices
A matrix is a bi-dimensional collection of data:
  

```{r}    
a <- matrix(1:12, nrow=3, ncol=4)           # define a matrix with 3 rows and 4 columns
a
```


```{r}
dim(a)                                      # return matrix dimensions (rows,columns)
```

```{r}
a <- matrix(1:8, nrow=4, ncol=4)            # create a matrix with 4 rows and 4 columns
a
```



### Arrays
They are similar to the matrices although they can have 2 o more dimensions.

```{r}
z <- array(1:24, dim=c(2,3,4))
z
```

### Factors
Factors are vectors that contain categorical information useful to group the values of other vectors of the same size. Let’s see an example:

```{r}  
bv <- c(0.92,0.97,0.87,0.91,0.92,1.04,0.91,0.94,0.96,
            +         0.90,0.96,0.86,0.85)          # (B-V) colours from 13 galaxies
```

If additional information is available (for instance, the morphological type of the galaxies) we can create a factor containing the galaxy types:
  

```{r}  
morfo <- c("Sab","E","Sab","S0","E","E","S0","S0","E","Sab","E","Sab","S0") # morphological info (same size)
length(morfo)                                       # ensure vector is the same size
```

```{r}  
fmorfo <- factor(morfo)                             # create factor with 'factor()'
fmorfo
```


```{r}  
levels(fmorfo)                                      # show factor levels
```

We could use this additional information to perform an statistical analysis segregating the data according to these types. This will be covered lately in the Functions section.

### Lists
Lists are ordered collections of objects, where the elements can be of a different type (a list can be a combination of matrices, vectors, other lists, etc.) They are created using the list() function:
  
```{r}  
gal <- list(name="NGC3379", morf="E", T.RC3=-5, colours=c(0.53,0.96))
gal
```


```

gal$<Tab>                   # pressing Tab key after '$', the elements of 'gal' are shown
#    gal$name     gal$morf     gal$T.RC3    gal$colours
```

```{r}
length(gal)                 # check how many elements 'gal' has
```

```{r}
names(gal)                  # return element names
```

```{r}
gal$radio <- TRUE                           # add a boolean element
gal$redshift <- 0.002922                    # add a numeric element
```

```{r} 
names(gal)                                  # return element names
```


Lists can be concatenated to generate bigger lists. If we have list1, list2, list3, we can create a unique list which is the result of the union of these three lists:

``` 
list123 <- c(list1, list2, list3)
```
As the elements in a list can be R objects of a different type:
  
Lists are extremely versatile since they can store every type of information (good)
Lists can be converted in objects with a rather complex structure (bad). A list can contain several elements which are vectors of different length, which is similar to having a table where the columns have a different number of rows.
The ideal situation is to take advantage of the list versatility but preventing them from growing with a very complex structure. This is why R has defined a new type of data which fulfils both requirements: a Data Frame.

### Data Frames (Tables)

A Data Frame is an special type of list very useful for the statistical work. There are some restrictions to guarantee that they can be used for this statistical purpose.

Among other restrictions, a Data Frame must verify that:
* List components must be vectors (numeric, character or logical vectors), factors, numeric matrices or other data frames.
* Vectors, which are the variables in the data frame, must be of the same length.

### Warning

In a data frame, character vectors are automatically converted into factors, and the number of levels can be determined as the number of different values in such a vector. This default behaviour can be modified with the options(stringsAsFactors = FALSE) command.
Basically, in a Data Frame all the information is displayed as a table where the columns have the same number of rows and can contain different type objects (numbers, characters, ...).

Data Frames can be created using the data.frame() function. Let’s see how to define a data frame with two elements, a numeric vector and a character vector (note that both must be same length vectors):
  
```{r}
options(stringsAsFactors = FALSE)
df <- data.frame(numbers=c(10,20,30,40),text=c("a","b","c","a"))
df
```

```{r}
df$text                                     # character vector not converted to a factor
```


```{r}
options(stringsAsFactors = TRUE)            # default
df <- data.frame(numbers=c(10,20,30,40),text=c("a","b","c","a"))
df$text
```


```{r}
df$numbers                               # numeric vector of length = 4
```

``` {r }
mode(df)                                    # storage mode of the object

```

```{r}
typeof(df)                                  # (internal) storage mode of the object
```

```{r}
class(df)                                   # object class
```


However the most common way of defining a data frame is reading the data stored in a file. We will see later how to do it using read.table() function.

#### Factors and Tables

It is frequently useful (for instance, for table creation) to be able to generate factors from a numeric continuum variable. To do so, we can use the cut command. Its parameter breaks defines how the data are divided. If breaks is a number, this is used as the number of (same length) intervals:

```{r} 
bv <- c(0.92,0.97,0.87,0.91,0.92,1.04,0.91,0.94,0.96,0.90,0.96,0.86,0.85)   # (B-V) colors from 13 galaxies
fbv <- cut(bv,breaks=3)                     # divide 'bv' in 3 equal-length intervals
fbv                                         # show in which interval every galaxy is
```


```{r}
table(fbv)                                  # generate a table with the 3 intervals
```


If breaks is a vector, its values are used as the limits of the intervals:
  
```{r}  
ffbv <- cut(bv,breaks=c(0.80,0.90,1.00,1.10))
table(ffbv)
ffbv
```

If we want just an approximate number of intervals, but with equally spaced round values, we can use the pretty() function (that not always returns the specified number of intervals!):

```{r}  
fffbv <- cut(bv,pretty(bv,3))               # ask for 3 'pretty' intervals
table(fffbv)                                # return 4 intervals
fffbv
```

We can also use a quantile division:
  
```{r}
ffffbv <- cut(bv,quantile(bv,(0:4)/4))      # ask for the 4 quantiles
table(ffffbv)
ffffbv
```

```{r}
heights <- c(1.64,1.76,1.79,1.65,1.68,1.65,1.86,1.82,1.73,1.75,1.59,1.87,1.73,1.57,1.63,1.71,1.68,1.73,1.53,1.82)
weights <- c(64,77,82,62,71,72,85,68,72,75,81,88,72,71,74,69,81,67,65,73)
ages    <- c(12,34,23,53,23,12,53,38,83,28,28,58,38,63,72,44,33,27,32,38)
```

For each one of these variables we can generate factors:

```{r}  
fheights <- cut(heights,c(1.50,1.60,1.70,1.80,1.90))        # factor for 'heights'
fweights <- cut(weights,c(60,70,80,90))                     # factor for 'weights'
fages    <- cut(ages,seq(10,90,10))                         # factor for 'ages'
```
Table generation is now straightforward using these factors. We can, for instance, generate bi-dimensional tables:

```{r}  
ta <- table(fheights, fweights)                     # table for 'heights' vs. 'weights'
ta
fweights
```


Marginal frequencies can also be included:

```{r}
addmargins(ta)
fweights
```

fheights    (60,70] (70,80] (80,90] Sum
(1.5,1.6]       1       1       1   3
(1.6,1.7]       2       3       1   6
(1.7,1.8]       2       4       1   7
(1.8,1.9]       1       1       2   4
Sum             6       9       5  20
Or we can work with the relative frequencies;

```{r}
tta <- prop.table(ta)
addmargins(tta)
fweights
```




### Exercises


1. Create a vector named v which contains 10 random integer values between -100 and +100 (?sample).
2. Create a two-dimensional 5?5 array named a comprised of sequence of even integers greater than 25. 

   * Create a list named s containing sequence of 20 capital letters, starting with 'C' (hint ?LETTERS) 

3. Create a list named l and put all previously created objects in it. Name them a, b and c respectively. 
How many elements are there in the list? Show the structure of the list. Count all elements recursively.

4. Without running commands in R, answer the following questions:

* what is the result of l[[3]]?
* How would you access random-th letter in the list element c?
* If you convert list l to a vector, what will be the type of it's elements?
* Can this list be converted to an array? What will be the data type of elements in array?
* Check the results with R.

5. Remove letters from the list l. Convert the list l to a vector and check its class. Compare it with the result from exercise 4, question #3.
6. Find the difference between elements in l[["a"]] and l[["b"]]. Find the intersection between them. Is there number 33 in their union?
7. Create 5?5 matrix named m and fill it with random numeric values rounded to two decimal places, ranging from 1.00 to 100.00.
8. Answer the following question without running R command, then check the result.
   What will be the class of data structure if you convert matrix m to:
   * vector
   * list
   * data frame
   * array?
   
