
### Exercises


1. Create a vector named v which contains 10 random integer values between -100 and +100.

```{r}
####################
#                  #
#    Exercise 1    #
#                  #
####################

v <- sample(-100:100, 10, replace=TRUE)

```

2. Create a two-dimensional 5?5 array named a comprised of sequence of even integers greater than 25.

  Create a list named s containing sequence of 20 capital letters, starting with 'C'.
   
```{r}
####################
#                  #
#    Exercise 2    #
#                  #
####################
a <- array(seq(from = 26, length.out = 25, by = 2), c(5, 5))
s <- LETTERS[match("C", LETTERS):(match("C", LETTERS)+19)]

```

3. Create a list named l and put all previously created objects in it. Name them a, b and c respectively. 
How many elements are there in the list? Show the structure of the list. Count all elements recursively.

```{r }
####################
#                  #
#    Exercise 3    #
#                  #
####################

l <- list(a = v, b = a, c = s)
length(l)
str(l)
length(unlist(l))

```

4. Without running commands in R, answer the following questions:

* what is the result of l[[3]]?
* How would you access random-th letter in the list element c?
* If you convert list l to a vector, what will be the type of it's elements?
* Can this list be converted to an array? What will be the data type of elements in array?
* Check the results with R.

```{r }
####################
#                  #
#    Exercise 4    #
#                  #
####################

l[[3]]
##  [1] "C" "D" "E" "F" "G" "H" "I" "J" "K" "L" "M" "N" "O" "P" "Q" "R" "S"
## [18] "T" "U" "V"
l[[3]][sample(1:length(l[[3]]), 1)]
## [1] "O"
class(unlist(l))
## [1] "character"
x <- array(l)
class(x[1])
## [1] "list"
```

5. Remove letters from the list l. Convert the list l to a vector and check its class. Compare it with the result from exercise 4, question #3.

```{r }
####################
#                  #
#    Exercise 5    #
#                  #
####################

l$c <- NULL
class(unlist(l))
## [1] "numeric"

```

6. Find the difference between elements in l[["a"]] and l[["b"]]. Find the intersection between them. Is there number 33 in their union?

```{r}
####################
#                  #
#    Exercise 6    #
#                  #
####################

setdiff(l$a, l$b)
## [1] -83 -44  71 -54 -17 -40 -76  22
intersect(l$a, l$b)
## [1] 72 58
33 %in% union(l$a, l$b)
## [1] FALSE
```

7. Create 5?5 matrix named m and fill it with random numeric values rounded to two decimal places, ranging from 1.00 to 100.00.

```{r }
####################
#                  #
#    Exercise 7    #
#                  #
####################

m <- matrix(data = round(runif(5*5, 0.99, 100.00), 2), nrow = 5)

```

8. Answer the following question without running R command, then check the result.
   What will be the class of data structure if you convert matrix m to:
   * vector
   * list
   * data frame
   * array?
   
```{r }
####################
#                  #
#    Exercise 8    #
#                  #
####################

class(as.vector(m))
## [1] "numeric"
class(as.list(m))
## [1] "list"
class(as.data.frame(m))
## [1] "data.frame"
class(as.array(m))
## [1] "matrix"

```
   
